---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# predictdiabetes

<!-- badges: start -->...
[![R-CMD-check](https://github.com/DSCI-310-2025/predictdiabetes/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/DSCI-310-2025/predictdiabetes/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/DSCI-310-2025/predictdiabetes/graph/badge.svg)](https://app.codecov.io/gh/DSCI-310-2025/predictdiabetes)
<!-- badges: end -->

## Overview

The goal of `predictdiabetes` is to provide functions that are useful for developing logistic regression models. This package was orginally designed for a logistic regression prediction model of diabetes risk based on BRFSS health indicators, but the package is designed to be applicable to other models.

## Functions

- `na_count_type`: Given an input data frame, check for the number of missing values (NA), the number of distinct values, and the data type of each variable.
- `category_target`: Given an input dataframe with a categorical variable, return the number and proportion of instances with each category value, in alphabetical (for characters and logical) or numerical (for integers) order
- `categorical_bars`: Create a list of bar plots for each categorical variable in a data frame.
- `info_gain`: Calculate and sort information gain for feature selection.
- `plots_grid`: Combine a list or mulitple lists of `ggplot2` objects into a single grid layout using `patchwork`. The number of columns in the grid can be specified with `num_cols`.
- `cramer_chi_results`: Runs chi-squared tests and calculates Cramer's V independently for each variable in a given data frame.

- `lr_pipeline`: Trains and fits a logistic regression model and cross-validates for optimal  hyperparameter values.

- `roc_plot`: Creates an ROC curve plot and saves it to the specified file path.
- `cm_plot`: Creates a confusion matrix plot and saves it to the specified file path.
- `coeff_plot`: Plot LASSO classification model coefficients.

## Installation

You can install the development version of predictdiabetes from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("DSCI-310-2025/predictdiabetes")
```

## Example

Begin with some EDA by checking the number of NAs, the number of distinct values and the data type in each column.

```{r example}
library(predictdiabetes)

na_count_type(mtcars)
#>                   mpg      cyl      disp     hp       drat     wt      
#> NA_Count          "0"      "0"      "0"      "0"      "0"      "0"     
#> Distinct_Count    "25"     "3"      "27"     "22"     "22"     "29"    
#> Current_Data_Type "double" "double" "double" "double" "double" "double"
#>                   qsec     vs       am       gear     carb    
#> NA_Count          "0"      "0"      "0"      "0"      "0"     
#> Distinct_Count    "30"     "2"      "2"      "3"      "6"     
#> Current_Data_Type "double" "double" "double" "double" "double"
```

When building the model, you will likely want to understand the proportions of the binary variable of interest within each category for categorical variables. `predictdiabetes::categorical_bars()` can be used to provide plots for every provided categorical variable.

```{r categorical_bars}
# Generate bar plots for categorical variables from the mtcars dataset
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$am   <- as.factor(mtcars$am)

bar_plots <- categorical_bars(
  data_frame = mtcars,
  cat_vars = c("cyl", "gear"),
  target_col = "am",
  title_size = 15,
  axis_size = 20
)

## Display one of the bar plots, e.g., for cyl:
(bar_plots[["cyl"]])
```

After building the LASSO model, ROC plots would be beneficial to evaluate how your model would perform.

```{r cm_plot}
set.seed(42)

conf_matrix_df <- data.frame(
  Prediction = c("Positive", "Negative", "Positive", "Negative"),
  Truth = c("Positive", "Positive", "Negative", "Negative"),
  Freq = c(50, 10, 5, 100)
)
plot <- cm_plot(conf_matrix_df, tempfile(fileext = ".png"))

plot(plot)
```

